Quantifying and Modeling Global Water Accessibility including Water Typology

### 2021-01-25 docs/manuscript/manuscript.tex

\citet{WHOdrinkingwater} article gives an ambitious plan to reach universal drinking water by the year of 2030 that is
affordable and safe followed by the Sustainable Development Goal six goals. The WHO/UNICEF Joint Monitoring Program for
Water Supply and Sanitation (JMP) have been globally monitoring progress on drinking water and sanitation since 1990. The
article outlines different plans with consistent traits to determine what is deemed to be safe and affordable drinking
water and good sanitation. Our study can highlight which modes of transportation of safe and affordable drinking water
can develop why one's community may excel in the improvement of safe and affordable drinking water and
sanitation. Drinking, sanitation, and hand washing ladder explains the qualitative determination can indirectly help
what is a good source of transporting safe and drinking water.

\citet{fuller2016tracking} highlights the global accessibility to safe drinking water and sanitation and concludes that
it has been improving well during the Millennium Development Goal Period. This topic will benefit our research to
understand what countries are doing for the mode of transportation of drinking water and sanitation which may have
contributed to the improvements. \citet{fuller2016tracking} chose some countries and studied the overall countries
progress which in his methodology revealed sigmoidal trends. Our study can reveal if certain methods of transportation
of drinking water and sanitation may have influenced the effectiveness of the improved drinking water met by MDG
standards.

@@

Next, there were many missing values from the data. Therefore, we had to do a simple missing value analysis. We determined to avoid using water, transportation, or wealth variables which had greater than 50\% missing values. Ultimately, simplified data frame was 76 countries by 28 variables. After, two different tables are generated where one focused on the water accessibility variables and a explanatory variables. We selected 11 water accessibility variables based on the sources available such as location, time, and methodology \citep{price2019difference}. The water accessibility variable was organized
in the following table to allow for quick visualization for \autoref{tab:summary-wa}. The other variables were used as an explanatory variables for determining the water accessibility predictors in \autoref{tab:summary-var}. 
\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Definition \\\midrule
    \texttt{ptap} & Households using a public tap/standpipe
 & Percentage of households whose main source of drinking water is a public tap/standpipe.
 \\\midrule
 \texttt{bore} & Households using a tube well/borehole & Percentage of households whose main source of drinking water is a tube-well/borehole.
\\\midrule
\texttt{pspr} & Households using a protected spring
& Percentage of households whose main source of drinking water is a protected spring.
\\\midrule
\texttt{uspr} & Households using an unprotected spring
& Percentage of households whose main source of drinking water is an unprotected spring.
\\\midrule
\texttt{pwel} & Households using a protected well
& Percentage of households whose main source of drinking water is a protected well.
\\\midrule
\texttt{uwel} & Households using an unprotected well water
&Percentage of households whose main source of drinking water is an unprotected well.
\\\midrule
\texttt{surw} & Households using surface water
& Percentage of households whose main source of drinking water is surface water.
\\\midrule
\texttt{tles} & Households with water 30 minutes or less away round trip
&Percentage of households with water 30 minutes or less away round trip.
\\\midrule 
\texttt{tmor} & Households with water more than 30 minutes away round trip & Percentage of households with water more than 30 minutes away round trip.
\\\midrule
\texttt{rain} & Households using rainwater
& Percentage of households whose main source of drinking water is rainwater.
\\\midrule
\texttt{phom} & Households using water piped into dwelling
&Percentage of households whose main source of drinking water is water piped into the dwelling.
\\\midrule

& \\\bottomrule
  \end{tabular}
  \caption{Summary of water accessibility variables (national averages at household level)}
  \label{tab:summary-wa}
\end{table}

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Definition \\\midrule
    \texttt{imws} & Improved water source
 & Drinking water is an improved source.
 \\\midrule
  \texttt{pipy} & Water piped into yard/plot
& Drinking water is water piped into the yard or plot.
\\\midrule
 \texttt{truc} & Tanker truck & Drinking water is tanker truck.
\\\midrule
 \texttt{ctan} & Cart with tank & Drinking water is cart with tank.
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john.
\\\midrule
 \texttt{uiws} & Unimproved water source & Drinking water is an unimproved source.
\\\midrule
 \texttt{othw} & Other water source & Drinking water is other.
\\\midrule
 \texttt{iwsp} & Improved water source on the premises & The de jure population living in households with an improved water source on the premises.
\\\midrule
 \texttt{basw} & Basic water service
 & Defined as an improved water source with either water on the premises or round-trip collection time is 30 minutes or less.
\\\midrule
 \texttt{liws} & Households with limited water service
 & Percentage of households with limited water service, defined as an improved water source with round-trip collection time greater than 30 minutes.
\\\midrule
 \texttt{prem} & Households with water on the premises
 & Percentage of households with water on the premises.
\\\midrule
 \texttt{bicy} & Households possessing a bicycle
 & Percentage of households possessing a bicycle.
\\\midrule
 \texttt{acar} & Households possessing an animal drawn cart
 & Percentage of households possessing an animal drawn cart.
\\\midrule
 \texttt{mcyc} & Households possessing a motorcycle
 & Percentage of households possessing a motorcycle.
\\\midrule
 \texttt{pcar} & Households possessing a private car
 & Percentage of households possessing a private car.
\\\midrule
 \texttt{boat} & Households possessing a boat with a motor
 & Percentage of households possessing a boat with a motor.
\\\midrule
 \texttt{wigc} & Wealth index Gini coefficient
 & The Gini coefficient indicates the level of concentration of wealth in the country.
\\\midrule
 & \\\bottomrule
  \end{tabular}
  \caption{Summary of explanatory variables (National Household usage Averages measured as percentage for Main Source)}
  \label{tab:summary-var}
\end{table}
 \subsection*{Summary of variables}
\hl{Here, you show and discuss the data distributions (e.g. histograms, correlation plots, etc)}

@@

\subsection{Water accessibility typologies}
\hl{Show the typology/cluster results and discuss.}
NBClust statistic analysis developed and determined four distinct groups. First cluster group composed of countries near the equator or in the Sub-Sahara Regions. The analysis of the second group represents countries considered developed or recently industrialized. Third cluster group represents most of the countries in the central Africa and may have similar factors as the first cluster group. The fourth cluster group represents most of countries in the middle east or predominately non-African countries. There is a great internally heterogeneous clusters for the 76 distinct countries. The cluster analysis' dendrogram is shown in \autoref{fig:dendrogram}, and the corresponding NBClust curve yielding the optimal four clusters is shown in \autoref{fig:optimal-clusters}, \autoref{fig:optimal}. \autoref{}

%   Table 3 shows the resulting five water and sanitation country clusters. Fig. 3 shows the cluster country membership on a world map. The cluster indicator variable means are summarized in Table 4. The Dunn and Davies-BouldinIndices for our clusters, as well as the World Bank, WHO, and MDGgroupings are shown in Table 5. Our clusters have both the largest Dunn index and smallest Davies-Bouldin index, indicating that they represent a superior grouping along with the chosen relevant indicators.

@@

\subsection{Model results}

>>>>

\citet{WHOdrinkingwater} article gives an ambitious plan to reach universal drinking water by the year 2030 that is
affordable and safe followed by the Sustainable Development Goal six goals. The WHO/UNICEF Joint Monitoring Program for
Water Supply and Sanitation (JMP) have been globally monitoring progress on drinking water and sanitation since 1990. \citet{WHOdrinkingwater} outlines different plans with consistent traits to determine what is deemed to be safe and affordable drinking
water and good sanitation.

\citet{fuller2016tracking} analyzes the global accessibility to safe drinking water and sanitation and concludes that
it has been improving well during the Millennium Development Goal Period. \citet{fuller2016tracking} chose some countries and studied the overall countries
progress which in his methodology revealed sigmoidal trends. Our study can reveal if certain methods of transportation
of drinking water and sanitation may have influenced the effectiveness of the improved drinking water met by MDGs
standards.

@@

Next, there were many missing values from the data. Therefore, we had to do a simple missing value analysis. We determined to avoid using water, transportation, or wealth variables which had greater than 50\% missing values. Ultimately, simplified data frame was 76 countries by 28 variables. After, two different tables are generated where one focused on the water accessibility variables and a explanatory variables. We selected 11 water accessibility variables based on the sources available such as location, time, and methodology \citep{price2019difference}. The water accessibility variable was organized
in the following table to allow for quick visualization for \autoref{tab:summary-wa}. The other variables were used as an explanatory variables for determining the water accessibility predictors in \autoref{tab:summary-var}. 
\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Definition \\\midrule
    \texttt{ptap} &  Public tap/standpipe
 & 
 \\\midrule
 \texttt{bore} & Tube well/borehole & 
\\\midrule
\texttt{pspr} & Protected spring
& 
\\\midrule
\texttt{uspr} & Unprotected spring
& 
\\\midrule
\texttt{pwel} & Protected well
& 
\\\midrule
\texttt{uwel} & Unprotected well water
&
\\\midrule
\texttt{surw} & Surface water
& 
\\\midrule
\texttt{tles} & Water 30 minutes or less away round trip
&
\\\midrule 
\texttt{tmor} & Water more than 30 minutes away round trip & 
\\\midrule
\texttt{rain} & Rainwater
& 
\\\midrule
\texttt{phom} & Piped into dwelling
&
\\\midrule

& \\\bottomrule
  \end{tabular}
  \caption{Summary of water accessibility variables (National Household usage Averages measured as percentage for Main Drinking Source)}
  \label{tab:summary-wa}
\end{table}

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Definition \\\midrule
    \texttt{imws} & Improved water source
 & 
 \\\midrule
  \texttt{pipy} & Water piped into yard/plot
&
\\\midrule
 \texttt{truc} & Tanker truck & 
\\\midrule
 \texttt{ctan} & Cart with tank & 
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john.
\\\midrule
 \texttt{uiws} & Unimproved water source & 
\\\midrule
 \texttt{othw} & Other water source & 
\\\midrule
 \texttt{iwsp} & Improved water source on the premises & The de jure population living in households with an improved water source on the premises.
\\\midrule
 \texttt{basw} & Basic water service
 & Defined as an improved water source with either water on the premises or round-trip collection time is 30 minutes or less.
\\\midrule
 \texttt{liws} & Limited water service
 & Limited water service, defined as an improved water source with round-trip collection time greater than 30 minutes.
\\\midrule
 \texttt{prem} & Water on the premises
 & 
\\\midrule
 \texttt{bicy} & Bicycle
 &
\\\midrule
 \texttt{acar} & Animal drawn cart
 & 
\\\midrule
 \texttt{mcyc} & Motorcycle
 & 
\\\midrule
 \texttt{pcar} & Private car
 & 
\\\midrule
 \texttt{boat} & Boat with a motor
 & 
\\\midrule
 \texttt{wigc} & Wealth index Gini coefficient
 & The Gini coefficient indicates the level of concentration of wealth in the country.
\\\midrule
 & \\\bottomrule
  \end{tabular}
  \caption{Summary of explanatory variables (National Household usage Averages measured as percentage for Main Drinking Source)}
  \label{tab:summary-var}
\end{table}
 \subsection*{Summary of variables}
\hl{Here, you show and discuss the data distributions (e.g. histograms, correlation plots, etc)}

@@

\subsection{Water accessibility typologies}

\hl{Show the typology/cluster results and discuss.}
NBClust statistic analysis developed and determined four distinct groups. The first cluster group is composed of countries near the equator or in the Sub-Sahara Regions. The analysis of the second group represents countries considered developed or recently industrialized. The third cluster group represents most of the countries in central Africa and may have similar factors as the first cluster group. The fourth cluster group represents most of the countries in the middle east or predominately non-African countries. There are great internally heterogeneous clusters for the 76 distinct countries. The cluster analysis' dendrogram is shown in \autoref{fig:dendrogram}, and the corresponding NBClust curve yielding the optimal four clusters is shown in \autoref{fig:optimal-clusters}, \autoref{fig:optimal}. \autoref{tab:water-cluster} shows the resulting four water accessibility country clusters. The cluster's water accessibility variables are summarized in \autoref{tab:summary-wa}. Through the optimal analysis of the clusters, which indicated the superior grouping of four cuts along with the relevant indicators.

%    Fig. 3 shows the cluster country membership on a world map. 
\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} }\toprule
    \bf Clusters & \bf Country Members \\\midrule
    \texttt{A (n=22)} & Afghanistan, Bangladesh, Benin, Burkina Faso, Cameroon, Chad, Ivory Coast, Ghana, Guinea, India, Liberia, Malawi, Mali, Myanmar, Nepal, Niger, Nigeria, Pakistan, Togo, Uganda, Zambia, Zimbabwe  \\\midrule
    \texttt{B (n=14)} & Albania, Armenia, Brazil, Colombia, Egypt, Honduras, Jordan, Kazakhstan, Morocco, Peru, South Africa, Turkey, Ukraine, Uzbekistan  \\\midrule
    \texttt{C (n=20)} & Angola, Burundi, Central African Republic, Congo, Congo Democratic Republic, Eritrea, Eswatini, Ethiopia, Gambia, Haiti, Kenya, Lesotho, Madagascar, Mauritania, Mozambique, Papua New Guinea, Rwanda, Sao Tome and Principe, Sierra Leone, Tanzania  \\\midrule
     \texttt{D (n=22)} & Azerbaijan, Bolivia, Cambodia, Comoros, Dominican Republic, Gabon, Guatemala, Guyana, Indonesia, Kyrgyz Republic, Maldives, Moldova, Namibia, Nicaragua, Paraguay, Philippines, Senegal, Tajikistan, Timor-Leste, Turkmenistan, Vietnam, Yemen  \\\bottomrule
  \end{tabular}
    \caption{Water accessibility country clusters.)}
  \label{tab:water-cluster}
\end{table}

@@

\subsection{Model results}
Cluster A consists of 22 countries with great cluster typology factor for

%Cluster 1 consists of 33 developed and recently industrializedcountries from North America, Europe, and the wealthier Asiancountries. Cluster 1 countries are characterized by the highest GDPsper capita and GE scores, relatively low levels of inequality as mea-sured by the Gini index, the highest expected years of education,high urbanization, and near-universal water and sanitation access.Example members are Canada, Germany, and Japan.
